Conceptual and methodological problems with bias detection and avoidance in natural language processing

### 2021-06-17 thesis/_book/_main.tex
small things

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  12pt,
]{book}
\usepackage{amsmath,amssymb}

@@

@@

@@

  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}

@@

% Use upquote if available, for straight quotes in verbatim environments

@@

\IfFileExists{microtype.sty}{% use microtype if available

@@

\@ifundefined{KOMAClassName}{% if non-KOMA class
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}

@@

  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[left=3.5cm, right=2.5cm, top=2.5cm, bottom=2.5cm, asymmetric, includeheadfoot]{geometry}

@@

\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}

@@

% Set default figure placement to htbp
\def\fps@figure{htbp}

@@

@@

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\ifluatex
  \usepackage{selnolig}  % disable illegal ligatures
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  % turn on hanging indent if param 1 is 1
  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
  % set entry spacing
  \ifnum #2 > 0
  \setlength{\parskip}{#2\baselineskip}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\title{Conceptual and methodological problems with bias detection and avoidance in natural language processing}
\author{Alicja Dobrzeniecka}
\date{2021-06-16}

@@

{
\setcounter{tocdepth}{5}
\tableofcontents
}
\setstretch{1.5}
\hypertarget{introduction}{%
\chapter{Introduction}\label{introduction}}

Placeholder

\hypertarget{cosine-similarity-and-bias-detection}{%
\chapter{Cosine similarity and bias detection}\label{cosine-similarity-and-bias-detection}}

\hypertarget{word-embeddings}{%
\section{Word embeddings}\label{word-embeddings}}

@@

\begin{longtable}[]{@{}llllll@{}}
\toprule
word & 1 & 2 & 3 & 4 & 5 \\
\midrule
\endhead
woman & 1 & 0 & 0 & 0 & 0 \\
man & 0 & 1 & 0 & 0 & 0 \\
girl & 0 & 0 & 1 & 0 & 0 \\
boy & 0 & 0 & 0 & 1 & 0 \\
monarch & 0 & 0 & 0 & 0 & 1 \\
\bottomrule
\end{longtable}

@@

\begin{longtable}[]{@{}lllll@{}}
\toprule
word & 1 & 2 & 3 & 4 \\
\midrule
\endhead
woman & 0.456 & 0.267 & 0.675 & 0.131 \\
man & 0.451 & 0.897 & 0.472 & 0.088 \\
girl & 0.604 & 0.262 & 0.414 & 0.706 \\
boy & 0.279 & 0.172 & 0.475 & 0.010 \\
monarch & 0.565 & 0.678 & 0.463 & 0.975 \\
\bottomrule
\end{longtable}

@@

\hypertarget{cosine-similarity-and-distance}{%
\section{Cosine similarity and distance}\label{cosine-similarity-and-distance}}

@@

\begin{align} \tag{Sim}
\mathsf{cosineSimilarity}(A,B) & = \frac{A \cdot B}{\vert \vert A \vert \vert \,\vert \vert B \vert \vert}
\end{align}
Cosine similarity is considered a proper tool for this operation as its result has a clear connection to geometry and at least for a low number of dimensions may be easily interpreted. Using this scale, one can compare vector similarities in a fairly clear manner. When the vectors are aligned perpendicularly to each other, their similarity equals 0 (which is the same as the cosine of 90 degrees). This tells us that the similarity between the vectors is small. As the angle between vectors decreases, cosine similarity approaches one, which stands for the greatest similarity.

@@

In what follows, it is important to distinguish between cosine similarity and cosine distance, defined as:
\begin{align} \tag{Sim}
\mathsf{cosineDistance}(A,B) &  = 1 - \mathsf{cosineSimilarity}(A,B)\\
 &  = 1 - \frac{A \cdot B}{\vert \vert A \vert \vert \,\vert \vert B \vert \vert} \nonumber
\end{align}

@@

\hypertarget{cosine-distance-in-a-one-class-bias-detection}{%
\section{Cosine distance in a one-class bias detection}\label{cosine-distance-in-a-one-class-bias-detection}}

\protect\hyperlink{ref-Bolukbasi2016Man}{Bolukbasi, Chang, Zou, Saligrama, \& Kalai} (\protect\hyperlink{ref-Bolukbasi2016Man}{2016}) define similarity between words as the outcome inner product of their normalized vectors.
They focus on examining what the geometry of word embedding is in regard to ``he'' and ``she'' words. In other words, whether the similarity between those concepts and other words reflects expected gender stereotypes. They test this hypothesis by investigating whether there is a connection between word embeddings representing certain professions and words referring to gender. They also evaluate whether automatically produced analogies between words reflect the stereotypes as well.

Here the gender bias of a word \(w\) is understood as its projection on the gender direction \(\vec{w} \cdot (\overrightarrow{he} - \overrightarrow{she})\) (the gender direction is the top principal compontent of ten gender pair difference vectors). The underlying idea is that no bias is present if non-explicitly gender words are in equal distance to both elements in all explicitly gender pairs. Given the (ideally) gender neutral words \(N\) and the gender direction \(g\) the direct direct gender bias is defined as the average distance of the words in \(N\) from \(g\) (\(c\) is a parameter determining how strict we want to be):
\begin{align}
\mathsf{directBias_c(N,g)} & = \frac{\sum_{w\in N}\vert \mathsf{cos}(\vec{w},g)\vert^c}{\vert N \vert }
\end{align}

@@

\begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/unnamed-chunk-1-1} \end{center}
\normalsize

@@

\hypertarget{cosine-distance-in-a-multi-class-bias-detection}{%
\section{Cosine distance in a multi-class bias detection}\label{cosine-distance-in-a-multi-class-bias-detection}}

\protect\hyperlink{ref-Manzini2019blackToCriminal}{Manzini, Lim, Tsvetkov, \& Black} (\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019}) present a different approach towards finding similarities between classes of words. The authors claim that texts available online are full of direct or indirect human stereotypes. As a result, word embeddings are prone to learn and maybe amplify those biases and propagate them further into AI models that are used for various applications. Cosine distance is used in the article as a measure to first argue for the existence of multi-class bias and then to show how through bias mitigation techniques the bias may be decreased.

They modify WEAT to a multi-class setting, introducing Mean Average Cosine similarity as a measure of bias (in fact, in the paper they report distances rather than similarities). Let \(T = \{t_1, \dots, t_k\}\) be a class of protected word embeddings, and let each \(A_j\in A\) be a set of attributes stereotypically associated with a protected word). Then:
\begin{align}
S(t_i, A_j) & = \frac{1}{\vert A_j\vert}\sum_{a\in A_j}\mathsf{cos}(t,a) \\
MAC(T,A) & = \frac{1}{\vert T \vert \,\vert A\vert}\sum_{t_i \in T }\sum_{A_j \in A} S(t_i,A_j)
\end{align}

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{religion }\OtherTok{\textless{}{-}} \FunctionTok{read.csv}\NormalTok{(}\StringTok{"../datasets/religionReddit.csv"}\NormalTok{)[}\SpecialCharTok{{-}}\DecValTok{1}\NormalTok{]}
\FunctionTok{colnames}\NormalTok{(religion) }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\StringTok{"protectedWord"}\NormalTok{,}\StringTok{"wordToCompare"}\NormalTok{,}\StringTok{"wordClass"}\NormalTok{,}
                        \StringTok{"cosineDistance"}\NormalTok{,}\StringTok{"cosineSimilarity"}\NormalTok{,}\StringTok{"connection"}\NormalTok{)}
\NormalTok{religion}\SpecialCharTok{$}\NormalTok{wordClass }\OtherTok{\textless{}{-}} \FunctionTok{as.factor}\NormalTok{(religion}\SpecialCharTok{$}\NormalTok{wordClass)}
\FunctionTok{levels}\NormalTok{(religion}\SpecialCharTok{$}\NormalTok{wordClass) }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\StringTok{"christian"}\NormalTok{,}\StringTok{"human"}\NormalTok{,}\StringTok{"jewish"}\NormalTok{,}\StringTok{"muslim"}\NormalTok{,}\StringTok{"neutral"}\NormalTok{)}
\FunctionTok{head}\NormalTok{(religion)  }\SpecialCharTok{\%\textgreater{}\%}  \FunctionTok{kable}\NormalTok{(}\AttributeTok{format =} \StringTok{"latex"}\NormalTok{,}\AttributeTok{booktabs=}\NormalTok{T,}
                      \AttributeTok{linesep =} \StringTok{""}\NormalTok{,  }\AttributeTok{escape =} \ConstantTok{FALSE}\NormalTok{, }
                      \AttributeTok{caption =} \StringTok{"Head of the religion dataset."}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
                      \FunctionTok{kable\_styling}\NormalTok{(}\AttributeTok{latex\_options=}\FunctionTok{c}\NormalTok{(}\StringTok{"scale\_down"}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

@@

\caption{\label{tab:religionTableHeadEarly}Head of the religion dataset.}
\centering
\resizebox{\linewidth}{!}{
\begin{tabular}[t]{lllrrl}
\toprule
protectedWord & wordToCompare & wordClass & cosineDistance & cosineSimilarity & connection\\
\midrule
judaism & violent & muslim & 0.7141939 & 0.2858061 & different\\
judaism & terrorist & muslim & 0.7461333 & 0.2538667 & different\\
judaism & dirty & muslim & 1.2002599 & -0.2002599 & different\\
judaism & uneducated & muslim & 0.7885469 & 0.2114531 & different\\
judaism & greedy & jewish & 1.0026172 & -0.0026172 & associated\\
judaism & cheap & jewish & 1.2323229 & -0.2323229 & associated\\
\bottomrule
\end{tabular}}
\end{table}
\normalsize

@@

\hypertarget{limitations-of-the-approach}{%
\section{Limitations of the approach}\label{limitations-of-the-approach}}

@@

Assuming for a moment that the value of multi-class cosine distance is correct, one may question the interpretation. \protect\hyperlink{ref-Manzini2019blackToCriminal}{Manzini, Lim, Tsvetkov, \& Black} (\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019}) summarize the averages of cosine distance per group (gender, race, religion). We would like to focus now on analyzing the values relating to religious biases. Here is the relevant fragment of table:

\begin{longtable}[]{@{}ll@{}}
\toprule
Religion Debiasing & MAC \\
\midrule
\endhead
Biased & 0.859 \\
Hard Debiased & 0.934 \\
Soft Debiased (\(\lambda\) = 0.2) & 0.894 \\
\bottomrule
\end{longtable}

@@

Using cosine similarity in high dimensions in word embeddings may also be prone to the curse of dimensionality. According to \protect\hyperlink{ref-Venkat2018Curse}{Venkat} (\protect\hyperlink{ref-Venkat2018Curse}{2018}) there are reasons to consider this phenomenon when searching for word similarities in higher dimensions. An experiment is conducted that aims at showing how the similarity values and variation change as the number of dimensions increases. The hypothesis made in the paper states that two things will happen as the number of dimensions increase. First, the effort required to measure cosine similarity will be greater, and two, the similarity between data will blur out and have less variation. The authors generate random points with increasing number of dimensions where each dimension of a data point is given a value between 0 and 1. Then they pick one vector at random from each dimension class and calculate the cosine similarity between the chosen vector and the rest of the data. Then they check how the variation of values changes as the number of dimensions increases. It seems like the more dimensions there are, the smaller the variance and therefore it is less obvious how to interpret the resulting cosine similarities. Maybe the scale should be adjusted to the number of dimensions and variance so that it still gives us sensible information about data. According to some researches the cosine similarity in high dimensions is not reliable enough to trust it as it may be the case that choosing words on random may result in getting similar values as when picking them consciously.

@@

\hypertarget{walkthrough-with-the-religion-dataset}{%
\chapter{Walkthrough with the religion dataset}\label{walkthrough-with-the-religion-dataset}}

Placeholder

\hypertarget{loading-and-understanding-the-dataset}{%
\section{Loading and understanding the dataset}\label{loading-and-understanding-the-dataset}}

\hypertarget{first-look-at-the-empirical-distributions}{%
\section{First look at the empirical distributions}\label{first-look-at-the-empirical-distributions}}

\hypertarget{looking-at-the-islam-related-words}{%
\section{Looking at the islam-related words}\label{looking-at-the-islam-related-words}}

\hypertarget{bayesian-model-structure-and-assumptions}{%
\section{Bayesian model structure and assumptions}\label{bayesian-model-structure-and-assumptions}}

\hypertarget{choosing-predictors}{%
\section{Choosing predictors}\label{choosing-predictors}}

\hypertarget{dataset-level-coefficients}{%
\section{Dataset-level coefficients}\label{dataset-level-coefficients}}

\hypertarget{model-structure-and-assumptions}{%
\section{Model structure and assumptions}\label{model-structure-and-assumptions}}

\hypertarget{protected-classes-in-reddit-and-google-embeddings}{%
\section{Protected classes in Reddit and Google embeddings}\label{protected-classes-in-reddit-and-google-embeddings}}

\hypertarget{dataset-level-coefficients-after-debiasing}{%
\section{Dataset-level coefficients after debiasing}\label{dataset-level-coefficients-after-debiasing}}

>>>>

\documentclass[12pt,]{book}

@@

\setstretch{1.5}
\usepackage{amssymb,amsmath}

@@

\usepackage{fixltx2e} % provides \textsubscript

@@

\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}

@@

  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
% use upquote if available, for straight quotes in verbatim environments

@@

% use microtype if available
\IfFileExists{microtype.sty}{%

@@

\PassOptionsToPackage{hyphens}{url} % url is loaded by hyperref
\usepackage[unicode=true]{hyperref}

@@

            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage[left=3.5cm, right=2.5cm, top=2.5cm, bottom=2.5cm, asymmetric,
includeheadfoot]{geometry}

@@

\usepackage{longtable,booktabs}
% Fix footnotes in tables (requires footnote package)
\IfFileExists{footnote.sty}{\usepackage{footnote}\makesavenoteenv{long table}}{}
\usepackage{graphicx,grffile}

@@

\setlength{\parskip}{6pt plus 2pt minus 1pt}
}

@@

% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}

% set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother

@@

\title{Conceptual and methodological problems with bias detection and avoidance
in natural language processing}
\author{Alicja Dobrzeniecka}
\date{2021-06-17}

@@

{
\setcounter{tocdepth}{5}
\tableofcontents
}
\chapter{Introduction}\label{introduction}

Natural language processing (NLP) is a subfield of computer science that
processes and analyzes language in text and speech with the use of
modern programming methods. It has practical applications in everyday
life as it concerns tasks such as email filters, smart assistants,
search results, language translations, text analytics and so on. Models
used to accomplish these tasks need a lot of data to learn from. This
data originates from human activities and historical recordings such as
texts, messages or speeches. It turns out that in the learning process
these models can learn implicit biases that reflect harmful
stereotypical thinking still present in modern societies. One can find
methods that aim at identifying and measuring hidden biases and/or try
to remove them by modifying the models. There are many different types
of models in NLP depending on a task that they are supposed to solve.
However, all of them need as an input words represented by means of
numbers and this is accomplished with word embedding models. The models
usually assign the values based on the context in which the words
appear. It means that the input data can have enormous influence on the
outcome. The biases seem to have their primary source in the way the
words are assigned the numerical values.

There is considerable amount of literature available on the topic of
bias detection and mitigation in NLP models. Bolukbasi, Chang, Zou,
Saligrama, \& Kalai (\protect\hyperlink{ref-Bolukbasi2016Man}{2016})
focus on gender biases that may be observable while investigating the
representation of job occupations and gender in terms of their assigned
numerical values. The authors apply cosine similarity measurement to
investigate the phenomenon where (the vectors corresponding to) words
related to jobs that are stereotypically associated with a given gender
are in fact in the model situated closer to this gender. They also use
analogy tasks to evaluate if the bias is present in the word embedding
model. They check analogies by comparing pairs of word vectors, for
example they search for the word complementing the puzzle: man is to
doctor as woman is to \ldots{}? First they subtract the word ``man''
from the word ``woman'' and then they search for the ranked list of
other words pairs that have similar vector difference. They also include
in the formula a threshold to ensure that the resulting pairs could not
be randomly picked.

However, as Nissim, Noord, \& Goot
(\protect\hyperlink{ref-Nissim2019Fair}{2019}) points out, there are
some limitations to this approach. In practice, most of analogies
implementations do not return any input words. This means that it does
not make sense to expect the algorithm to return the same profession for
both woman and man. Therefore, this method seems to be limited in terms
of bias detection. Other problems are related for example to the choice
of pairs and words that are used to detect the presence of
discrimination as it is often subjective and without proper
justification. Additionally the choice of the parameter used in
Bolukbasi et al. (\protect\hyperlink{ref-Bolukbasi2016Man}{2016})
formula to ensure that word pairs are not picked by random, is also not
justified and changing it drastically influences the results.

Islam, Bryson, \& Narayanan
(\protect\hyperlink{ref-Caliskan2017Semantics}{2016}) touch upon the
topic of biases regarding race and gender. They apply knowledge from
well-known psychological studies such as Implicit Association Test to
research the relation between human stereotypical thinking and model
learnt biases to discover close relationship between these two. For the
evaluation they use Word Embedding Association Test (WEAT) and the Word
Embedding Factual Association Test (WEFAT).

({\textbf{???}}) propose a novel way of using cosine similarity to
obtain information on assumed resemblance between words. They
investigate an approach that enables them to measure the bias for a
class (like gender, religion, race) and express the final result with a
single metric.

Islam et al. (\protect\hyperlink{ref-Caliskan2017Semantics}{2016}) refer
to Implicit Association Test (Greenwald et al., 1998) that measures the
strength of associations between concepts or stereotypes by calculating
the time of reaction for special tasks. Humans naturally exhibit some
biases which do not always cause social concern. One can imagine the
intuitive associations between for example insects and flowers, and the
feelings of pleasantness or unpleasantness. In general, people would
rather associate flowers with feeling pleasant than insects, and this
preference in a sense is a bias or prejudice in some direction. However,
this type of preference does not cause an uproar and is a rather morally
neutral case. This example is rather used to show that the methodology
used in IAT makes sense. Unfortunately, the further studies discover
other biases and prejudices that directly influence the quality of other
people's lives and therefore they should be taken care of.

One can find various definitions trying to capture what bias and
fairness actually are. With the choice of the definition, implications
into the real-life applications may change as well. Mehrabi, Morstatter,
Saxena, Lerman, \& Galstyan
(\protect\hyperlink{ref-Mehrabi2019Survey}{2019}) mark out that there
exist different types of biases such as historical bias, representation
bias or measurement bias (the list is long). This indicates how complex
the issue of bias is. Without the proper understanding and awareness of
the problem, people are prone to unconsciously sustain the bias.

Mehrabi et al. (\protect\hyperlink{ref-Mehrabi2019Survey}{2019}) also
distinguish different types of discrimination, some of them will be
briefly described. By protected attributes we mean those qualities,
traits or characteristics that one should not discriminate against.
Direct discrimination occurs when protected attributes of individuals
explicitly result in non-favorable outcomes toward them. Such
discrimination takes place for example when one uses an information
about someone's religion belief to evaluate whether they are competent
for the work they apply for. In contrast in indirect discrimination
individuals appear to be treated equally but anyway they end up being
treated unjustly due to the hidden effects of biases towards their
protected attributes. An example of such situation is when in a job
advert there is a requirement of 10 years experience instead of the list
of a specific type of experience and knowledge. This job advert can
discriminate indirectly against young people who can have equivalent
skills but not yet so much work experience. Systemic discrimination
takes place when policies, customs or behaviors that result from certain
culture or organizational structure lead to discrimination against some
groups of people. Finally, statistical discrimination consists in to
using average group statistics to judge person belonging to the group.

The topic of discrimination is entangled with another concept; fairness.
It is essential to grasp some concepts of fairness to take them into
consideration while designing implementation of some machine learning
model. In Mehrabi et al.
(\protect\hyperlink{ref-Mehrabi2019Survey}{2019}) one may notice that
depending on the context and application different definitions may be
applied.

The most common methods in the literature focus on comparing the
similarity between words from protected groups and those that are
considered to be stereotypical or harmful in some way. One can find in
this group methods such as euclidean distance or cosine similarity
(which is equivalent to dot product if the vectors are normalized).
There are also other ways to detect the effects of biases. One can
investigate the model performance on certain tasks and validate if the
model uses the gender or race information to complete the task.

The currently used methods (such as cosine similarity) make the
similarity values often aggregated in a way that may lead to hasty
conclusions. The averaging of values and the lack of uncertainty may
lead to the incomplete picture of the situation.

In the paper we indicate how current methods used to detect biases in
natural language models are limited, if we investigate them from the
perspective of Bayesian analysis.

Our research enhances the current way in which the bias detection is
performed to make sure that it is methodologically valid.

The key hypothesis is that greater understanding of data and bias
implications can be achieved when Bayesian methods are applied to issue.

\chapter{Cosine similarity and bias
detection}\label{cosine-similarity-and-bias-detection}

\section{Word embeddings}\label{word-embeddings}

@@

\begin{longtable}[]{@{}llllll@{}}
\toprule
word & 1 & 2 & 3 & 4 & 5\tabularnewline
\midrule
\endhead
woman & 1 & 0 & 0 & 0 & 0\tabularnewline
man & 0 & 1 & 0 & 0 & 0\tabularnewline
girl & 0 & 0 & 1 & 0 & 0\tabularnewline
boy & 0 & 0 & 0 & 1 & 0\tabularnewline
monarch & 0 & 0 & 0 & 0 & 1\tabularnewline
\bottomrule
\end{longtable}

@@

\begin{longtable}[]{@{}lllll@{}}
\toprule
word & 1 & 2 & 3 & 4\tabularnewline
\midrule
\endhead
woman & 0.456 & 0.267 & 0.675 & 0.131\tabularnewline
man & 0.451 & 0.897 & 0.472 & 0.088\tabularnewline
girl & 0.604 & 0.262 & 0.414 & 0.706\tabularnewline
boy & 0.279 & 0.172 & 0.475 & 0.010\tabularnewline
monarch & 0.565 & 0.678 & 0.463 & 0.975\tabularnewline
\bottomrule
\end{longtable}

@@

\section{Cosine similarity and
distance}\label{cosine-similarity-and-distance}

@@

\begin{align} \tag{Sim}
\mathsf{cosineSimilarity}(A,B) & = \frac{A \cdot B}{\vert \vert A \vert \vert \,\vert \vert B \vert \vert}
\end{align}

Cosine similarity is considered a proper tool for this operation as its
result has a clear connection to geometry and at least for a low number
of dimensions may be easily interpreted. Using this scale, one can
compare vector similarities in a fairly clear manner. When the vectors
are aligned perpendicularly to each other, their similarity equals 0
(which is the same as the cosine of 90 degrees). This tells us that the
similarity between the vectors is small. As the angle between vectors
decreases, cosine similarity approaches one, which stands for the
greatest similarity.

@@

In what follows, it is important to distinguish between cosine
similarity and cosine distance, defined as:

\begin{align} \tag{Sim}
\mathsf{cosineDistance}(A,B) &  = 1 - \mathsf{cosineSimilarity}(A,B)\\
 &  = 1 - \frac{A \cdot B}{\vert \vert A \vert \vert \,\vert \vert B \vert \vert} \nonumber
\end{align}

@@

\section{Cosine distance in a one-class bias
detection}\label{cosine-distance-in-a-one-class-bias-detection}

Bolukbasi et al. (\protect\hyperlink{ref-Bolukbasi2016Man}{2016}) define
similarity between words as the outcome inner product of their
normalized vectors. They focus on examining what the geometry of word
embedding is in regard to ``he'' and ``she'' words. In other words,
whether the similarity between those concepts and other words reflects
expected gender stereotypes. They test this hypothesis by investigating
whether there is a connection between word embeddings representing
certain professions and words referring to gender. They also evaluate
whether automatically produced analogies between words reflect the
stereotypes as well.

Here the gender bias of a word \(w\) is understood as its projection on
the gender direction
\(\vec{w} \cdot (\overrightarrow{he} - \overrightarrow{she})\) (the
gender direction is the top principal compontent of ten gender pair
difference vectors). The underlying idea is that no bias is present if
non-explicitly gender words are in equal distance to both elements in
all explicitly gender pairs. Given the (ideally) gender neutral words
\(N\) and the gender direction \(g\) the direct direct gender bias is
defined as the average distance of the words in \(N\) from \(g\) (\(c\)
is a parameter determining how strict we want to be):

\begin{align}
\mathsf{directBias_c(N,g)} & = \frac{\sum_{w\in N}\vert \mathsf{cos}(\vec{w},g)\vert^c}{\vert N \vert }
\end{align}

@@

\begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/unnamed-chunk-1-1} \end{center}

\normalsize

@@

\section{Cosine distance in a multi-class bias
detection}\label{cosine-distance-in-a-multi-class-bias-detection}

Manzini, Lim, Tsvetkov, \& Black
(\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019}) present a
different approach towards finding similarities between classes of
words. The authors claim that texts available online are full of direct
or indirect human stereotypes. As a result, word embeddings are prone to
learn and maybe amplify those biases and propagate them further into AI
models that are used for various applications. Cosine distance is used
in the article as a measure to first argue for the existence of
multi-class bias and then to show how through bias mitigation techniques
the bias may be decreased.

They modify WEAT to a multi-class setting, introducing Mean Average
Cosine similarity as a measure of bias (in fact, in the paper they
report distances rather than similarities). Let
\(T = \{t_1, \dots, t_k\}\) be a class of protected word embeddings, and
let each \(A_j\in A\) be a set of attributes stereotypically associated
with a protected word). Then:

\begin{align}
S(t_i, A_j) & = \frac{1}{\vert A_j\vert}\sum_{a\in A_j}\mathsf{cos}(t,a) \\
MAC(T,A) & = \frac{1}{\vert T \vert \,\vert A\vert}\sum_{t_i \in T }\sum_{A_j \in A} S(t_i,A_j)
\end{align}

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{religion <-}\StringTok{ }\KeywordTok{read.csv}\NormalTok{(}\StringTok{"../datasets/religionReddit.csv"}\NormalTok{)[}\OperatorTok{-}\DecValTok{1}\NormalTok{]}
\KeywordTok{colnames}\NormalTok{(religion) <-}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"protectedWord"}\NormalTok{,}\StringTok{"wordToCompare"}\NormalTok{,}\StringTok{"wordClass"}\NormalTok{,}
                        \StringTok{"cosineDistance"}\NormalTok{,}\StringTok{"cosineSimilarity"}\NormalTok{,}\StringTok{"connection"}\NormalTok{)}
\NormalTok{religion}\OperatorTok{$}\NormalTok{wordClass <-}\StringTok{ }\KeywordTok{as.factor}\NormalTok{(religion}\OperatorTok{$}\NormalTok{wordClass)}
\KeywordTok{levels}\NormalTok{(religion}\OperatorTok{$}\NormalTok{wordClass) <-}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"christian"}\NormalTok{,}\StringTok{"human"}\NormalTok{,}\StringTok{"jewish"}\NormalTok{,}\StringTok{"muslim"}\NormalTok{,}\StringTok{"neutral"}\NormalTok{)}
\KeywordTok{head}\NormalTok{(religion)  }\OperatorTok{%>%}\StringTok{  }\KeywordTok{kable}\NormalTok{(}\DataTypeTok{format =} \StringTok{"latex"}\NormalTok{,}\DataTypeTok{booktabs=}\NormalTok{T,}
                      \DataTypeTok{linesep =} \StringTok{""}\NormalTok{,  }\DataTypeTok{escape =} \OtherTok{FALSE}\NormalTok{, }
                      \DataTypeTok{caption =} \StringTok{"Head of the religion dataset."}\NormalTok{) }\OperatorTok{%>%}
\StringTok{                      }\KeywordTok{kable_styling}\NormalTok{(}\DataTypeTok{latex_options=}\KeywordTok{c}\NormalTok{(}\StringTok{"scale_down"}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

@@

\caption{\label{tab:religionTableHeadEarly}Head of the religion dataset.}
\centering
\resizebox{\linewidth}{!}{
\begin{tabular}[t]{lllrrl}
\toprule
protectedWord & wordToCompare & wordClass & cosineDistance & cosineSimilarity & connection\\
\midrule
judaism & violent & muslim & 0.7141939 & 0.2858061 & different\\
judaism & terrorist & muslim & 0.7461333 & 0.2538667 & different\\
judaism & dirty & muslim & 1.2002599 & -0.2002599 & different\\
judaism & uneducated & muslim & 0.7885469 & 0.2114531 & different\\
judaism & greedy & jewish & 1.0026172 & -0.0026172 & associated\\
judaism & cheap & jewish & 1.2323229 & -0.2323229 & associated\\
\bottomrule
\end{tabular}}
\end{table}

\normalsize

@@

\section{Limitations of the approach}\label{limitations-of-the-approach}

@@

Assuming for a moment that the value of multi-class cosine distance is
correct, one may question the interpretation. Manzini et al.
(\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019}) summarize the
averages of cosine distance per group (gender, race, religion). We would
like to focus now on analyzing the values relating to religious biases.
Here is the relevant fragment of table:

\begin{longtable}[]{@{}ll@{}}
\toprule
Religion Debiasing & MAC\tabularnewline
\midrule
\endhead
Biased & 0.859\tabularnewline
Hard Debiased & 0.934\tabularnewline
Soft Debiased (\(\lambda\) = 0.2) & 0.894\tabularnewline
\bottomrule
\end{longtable}

@@

Using cosine similarity in high dimensions in word embeddings may also
be prone to the curse of dimensionality. According to Venkat
(\protect\hyperlink{ref-Venkat2018Curse}{2018}) there are reasons to
consider this phenomenon when searching for word similarities in higher
dimensions. An experiment is conducted that aims at showing how the
similarity values and variation change as the number of dimensions
increases. The hypothesis made in the paper states that two things will
happen as the number of dimensions increase. First, the effort required
to measure cosine similarity will be greater, and two, the similarity
between data will blur out and have less variation. The authors generate
random points with increasing number of dimensions where each dimension
of a data point is given a value between 0 and 1. Then they pick one
vector at random from each dimension class and calculate the cosine
similarity between the chosen vector and the rest of the data. Then they
check how the variation of values changes as the number of dimensions
increases. It seems like the more dimensions there are, the smaller the
variance and therefore it is less obvious how to interpret the resulting
cosine similarities. Maybe the scale should be adjusted to the number of
dimensions and variance so that it still gives us sensible information
about data. According to some researches the cosine similarity in high
dimensions is not reliable enough to trust it as it may be the case that
choosing words on random may result in getting similar values as when
picking them consciously.

@@

\chapter{Walkthrough with the religion
dataset}\label{walkthrough-with-the-religion-dataset}

\section{Loading and understanding the
dataset}\label{loading-and-understanding-the-dataset}

We start with loading the libraries needed for the analysis.

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{library}\NormalTok{(ggplot2)}
\KeywordTok{library}\NormalTok{(ggthemes)}
\KeywordTok{library}\NormalTok{(rethinking)}
\KeywordTok{library}\NormalTok{(tidyverse)}
\KeywordTok{library}\NormalTok{(ggpubr)}
\KeywordTok{library}\NormalTok{(kableExtra)}
\KeywordTok{library}\NormalTok{(dplyr)}
\KeywordTok{library}\NormalTok{(ggExtra)}
\KeywordTok{library}\NormalTok{(cowplot)}
\end{Highlighting}
\end{Shaded}

\normalsize

We will use the choice of protected words and stereotypical predicates
used in REF. This is a decent point of departure, not only we want to
compare our method to that of REF, but also because this data format is
fairly general (as contrasted, say, with a set up for binary
stereotypes). Note also that the method we develop here can fairly
easily be run for different stereotypization patterns. Let's start with
explaining the method and its deployment using a dataset obtained for
the religion-related protected words.

Let's load, clean a bit and inspect the head of the religion dataset we
prepared. In order to obtain this dataset, we calculated the cosine
distance between each protected word and each word from both the
bias-related attribute groups, which were used in the original study,
and to neutral and human control attributes which we added as control
groups. For instance, for religion, the bias-related predicates (coming
from the original study in REF) include muslim bias attributes, jew bias
attributes, christian bias attributes (see a list in the APPENDIX).
